### 2016-04-18 true-grid/01_Abstract.tex
added a folder for SEMS

%!TEX root = 00_Paper_SEMS16.tex
\begin{abstract}
In the past years, we have seen several papers analyzing large spreadsheet corpora. Also, there is a growing body of research on the appliance of software engineering methods on spreadsheets. These studies focus to a large extent on formulas in spreadsheets. However, there are other interesting concepts in spreadsheets like Pivot Tables and charts. Users make use of these concepts to analyze and visualize large data sets. Despite the popularity of Pivot Tables and Charts, there is almost no research about their use. In this paper, we analyze two popular spreadsheet corpora: ENRON and EUSES on the use of Pivot Tables and charts.
\end{abstract}

### 2016-04-18 true-grid/paper_SEMS16.tex
added a folder for SEMS

\documentclass{llncs}
\usepackage{graphicx}
\newcommand{\todo}[1]{\textbf{#1}}

\begin{document}
\title{TrueGrid: Code in a Grid and Grids in the IDE}
\author{Felienne Hermans \and Tijs van der Storm}
\institute{TODO\\
	\email{f.f.j.hermans@tudelft.nl}\\
	\email{todo}} 
\maketitle
\begin{abstract}
Spreadsheet systems are live programming environments. Both the data and the code are right in front you, and if you edit either of them, the effects are immediately visible. Unfortunately, spreadsheets lack mechanisms for abstraction, such as classes, function definitions etc. Programming languages excel at abstraction, but most mainstream languages or integrated development environments (IDEs) do not support the interactive, live feedback loop of spreadsheets. As a result, exploring and testing of code is cumbersome and indirect.

In this paper we propose a method to bring both worlds closer together. We propose TrueGrid: an approach where spreadsheet cells can be programmed not with formulas but a full featured programming language (JavaScript).

We then describe TrueGrid within two different worlds: the spreadsheet world, where it adds the benefits of source code, including added abstractions, syntax highlighting, version control. We then describe TrueGrid within the IDE, where it brings the benefits of the liveness of spreadsheets which will support rapid application development.
\end{abstract}

\section{Introduction}
\subsection{Spreadsheets}
Spreadsheets are very popular tools for end-user programming \todo{cite proof}. Their formula language is easy to learn and their grid interface is inviting. Apart from this, spreadsheets have the interesting propoerty of being \emph{live} programming. First proposed as a design principle by Maloney and Smith \cite{maloney_directness_1995}, liveness indicates that a user interface is always active and reactive. According to \cite{maloney_directness_1995} in a live interface ``objects respond to user actions, animations run, layout happens, and information displays are updated continuously''. More recently live programming has found its way to the public eye, among others by Bret Victor in his talk `Inventing on Principle' \cite{Victor2012}. Figure \ref{fig:bret}, taken from Victor's talk, illustrates the idea of live programming: on the right, we have source code and on the left, we have the result of that code, in this case: a tree. Modifying the code will immediately affect the tree.

This liveness is one of the core features of most spreadsheet systems. When a users enters a formula and presses enter, they see the result, without any effort such as compilation. Liveness of spreadsheets powers a spreadsheet's flexibility, often praised as their key success factor.

\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{fig/bret.png}
  \caption{Live programming: on the right the source code and on the left its instantiation of the code which changes immediately when the code is updated, screenshot from~\cite{Victor2012}.}
  \label{fig:bret}
  \end{center}
\end{figure}

However, spreadsheets have a number of downsides too, for example the lack of possibilities to abstraction. Instead of creating objects and methods on them, spreadsheet users mimic abstraction by copy-pasting formulas \todo{references naar LIVE paper}. Secondly, the editor that Excel and other spreadsheet systems provide is very limited, for example lacking syntax highlighting and not inviting proper structuring of formulas.

\subsection{Code}
On the other hand, there is source code. Most existing programming languages do support abstraction, and modern editors help developers understand and structure their code. However, source code does not have liveness \todo{apart from a few recent initiative that have not been adopted mainstraim, Tijs, hedd gij dr paraat?}

%\subsubsection{Directness}
%Another benefit of spreadsheets is that their interface combines data, metadata and calculations together in one view, and provides the user with easy access to all. Just by clicking a formula, one can manipulate it. This is often called `directness': \emph{``the feeling that one is directly manipulating the object''} \cite{shneiderman_direct_1983}. From a cognitive perspective, directness in computing means \emph{``a small distance between a goal and the actions required of the user to achieve the goal''} \cite{burnett_visual_2001}.

%Maloney and Smith describe directness as the fact that a user can \emph{``can initiate the process of examining or changing the attributes, structure, and behavior of user interface components by pointing at their graphical representations directly, as opposed to navigating through an alternate representation.''}\cite{maloney_directness_1995}.

%This almost exactly describes the interface of a spreadsheet. Instead of navigating to a code behind, a class or an object, spreadsheet users have all ingredients: data, metadata and calculations, in one view, and can access them with one click.

\section{TrueGrid}
In this paper, we will describe an approach bridging the gap between programming and spreadsheets and describe ts implications in both worlds. The basic concept of TrueGrid is a grid which can be programmed in a fully featured programming language, where the grid and the code are juxtapositioned ed, like in a spreadsheet, as demonstrated in Figure \ref{fig:TG}. Furthermore, like a spreadsheet, TrueGrid is live, i.e. on a change of data or code, the grid is updated.

\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{fig/TG.png}
  \caption{A True Grid \todo{make ss with syntax highlighting, and add explanation}}
  \label{fig:TG}
  \end{center}
\end{figure}

In our example, we use JavaScript as a language, however, the idea itself is not limited to one programming language, in fact, one can even image a grid in which different programming languages live.

\section{TrueGrid for Spreadsheet Users}
As an implementation of TrueGrid for spredsheet users, we hypothesize that it will be embedded within an existing spreadsheet system, allowing users to use advanced features like PivotTables and charts on top of their TrueGrid. For spreadsheet users, using TrueGrid over spreadsheets presents several benefits. For example, a programming language enables the use of modern editors. In the above example we have used Ace \todo{add link} providing syntax highlighting and syntax errors for free. Furthermore, the textual form of the code allows for easy diffing, merging enabling more mature version control on grids. While learning a new programming language can be challenging, there are spreadsheet developers working with VBA now, which is a fully featured language, and we envision TrueGrid having a lower threshold as the liveness and juxtaposition will ease understanding.

\section{TrueGrid for Developers}
Like spreadsheet users, developers often work with (example) data when programming. In modern IDEs, it is not possible to load data and then program alongside it. This makes it difficult for quickly prototype simple data analysis projects. A TrueGrid embedded in the IDE

Using TrueGrid, a programmer you can inspect data and play with an initial implementation, before having to write interfaces or even unit tests. We envision this is a useful tool very early in the development process, when you have some data and only a vague idea of where the project is going.

For example, consider the situation where a developer has a list of number that you want to manipulate \todo{We need a better example here} Before she even wants to write tests, the developer might want to explore the dataset a bit, for example, by sorting, calculating the mean or outliers. These type of activities currently have to be done outside of the IDE, for example in Excel or using a scripting language like R. Using TrueGrid this activity can be performed within the IDE.

\section{Related Work}
\todo{tsja}

\bibliographystyle{splncs03}
\bibliography{references}
\end{document}

### 2016-04-19 true-grid/paper_SEMS16.tex
5 pages

@@
 \usepackage{graphicx}
+\usepackage{listings}
+\usepackage{color}
+\usepackage{array}
+\usepackage{url}
+
 \newcommand{\todo}[1]{\textbf{#1}}
 
+\definecolor{lightgray}{rgb}{.9,.9,.9}
+\definecolor{darkgray}{rgb}{.4,.4,.4}
+\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
+
+\lstdefinelanguage{JavaScript}{
+  keywords={typeof, new, true, false, catch, function, return, null, catch, switch, var, if, in, while, do, else, case, break},
+  keywordstyle=\color{blue}\bfseries,
+  ndkeywords={class, export, boolean, throw, implements, import, this},
+  ndkeywordstyle=\color{darkgray}\bfseries,
+  identifierstyle=\color{black},
+  sensitive=false,
+  comment=[l]{//},
+  morecomment=[s]{/*}{*/},
+  commentstyle=\color{purple}\ttfamily,
+  stringstyle=\color{red}\ttfamily,
+  morestring=[b]',
+  morestring=[b]"
+}
+
+\lstset{
+   language=JavaScript,
+   backgroundcolor=\color{white},
+   extendedchars=true,
+   basicstyle=\footnotesize\ttfamily,
+   showstringspaces=false,
+   showspaces=false,
+   numbers=none,
+   numberstyle=\footnotesize,
+   numbersep=9pt,
+   tabsize=2,
+   breaklines=true,
+   showtabs=false,
+   captionpos=b
+}
+
 \begin{document}
-\title{TrueGrid: Code in a Grid and Grids in the IDE}
+\title{TrueGrid: Code the Table, Tabulate the Data}
 \author{Felienne Hermans \and Tijs van der Storm}
-\institute{TODO\\
+\institute{TU Delft and CWI\\
 	\email{f.f.j.hermans@tudelft.nl}\\
-	\email{todo}} 
+	\email{storm@cwi.nl}} 
 \maketitle
@@
 
-In this paper we propose a method to bring both worlds closer together.
-We propose TrueGrid: an approach where spreadsheet cells can be programmed not with formulas but a full featured programming language (JavaScript).
-
-We then describe TrueGrid within two different worlds: the spreadsheet world, where it adds the benefits of source code, including added abstractions, syntax highlighting, version control.
-We then describe TrueGrid within the IDE, where it brings the benefits of the liveness of spreadsheets which will support rapid application development.
+In this paper we propose a method to bring both worlds closer together, by juxtaposing ordinary code and spreadsheet-like grids in the IDE, called TrueGrid.
+Using TrueGrid spreadsheet cells can be programmed using full featured programming language.
+Spreadsheet users then may enjoy benefits of source code, including added abstractions, syntax highlighting, version control, etc.
+On the other hand, programmers may leverage the grid for interactive exploring and testing of code.
+We illustrate these benefits using a prototype implementation of TrueGrid that runs in the browser an uses Javascript as a programming language.
 \end{abstract}
@@
 \section{Introduction}
-\subsection{Spreadsheets}
+%\subsection{Spreadsheets}
 Spreadsheets are very popular tools for end-user programming \todo{cite proof}.
 Their formula language is easy to learn and their grid interface is inviting.
-Apart from this, spreadsheets have the interesting propoerty of being \emph{live} programming.
-First proposed as a design principle by Maloney and Smith \cite{maloney_directness_1995}, liveness indicates that a user interface is always active and reactive.
-According to \cite{maloney_directness_1995} in a live interface ``objects respond to user actions, animations run, layout happens, and information displays are updated continuously''.
-More recently live programming has found its way to the public eye, among others by Bret Victor in his talk `Inventing on Principle' \cite{Victor2012}.
-Figure \ref{fig:bret}, taken from Victor's talk, illustrates the idea of live programming: on the right, we have source code and on the left, we have the result of that code, in this case: a tree.
-Modifying the code will immediately affect the tree.
+Apart from this, spreadsheets are \textit{live}: the  user interface reacts immediately to changes in input or code.
+Live programming helps bridging the gulf between code and behavior because the user receives immediate feedback on their actions~\cite{lieberman1995bridging}.
+More recently, live programming has found its way to a wider audience, for instance, by Bret Victor's influential talk \textit{Inventing on Principle}~\cite{Victor2012}.
+Figure \ref{fig:bret}, taken from Victor's talk, illustrates the idea of live programming: on the right, we have source code and on the left, we have the result of that code, in this case, a tree drawing.
+Modifying the code will immediately affect the visual representation of the tree.
 
 This liveness is one of the core features of most spreadsheet systems.
-When a users enters a formula and presses enter, they see the result, without any effort such as compilation.
-Liveness of spreadsheets powers a spreadsheet's flexibility, often praised as their key success factor.
-
+When a users enters a formula and presses enter, they see the result, without long edit-compile-run cycles.
+This live characteristic of spreadsheets is often praised as their key success factor~\cite{proof}.
 \begin{figure}
@@
 \end{figure} 
+However, spreadsheets have a number of downsides too.
+For instance, the lack of  abstraction mechanisms forces spreadsheet users to use copy-paste for reusing code.
+Although solutions to this problem have been researched~\cite{hermans_2015_19341,jones2003user}, they do not provide the power of full programming languages to spreadsheet users.
+Secondly, the way formulas are edited in a spreadsheet system like Excel has important drawbacks from the programming perspective.
+For instance, Excel's formula editor lacks even the basic editor services, such as  syntax highlighting and reference resolution.
+In general, the interface is not inviting to apply proper coding styles and practices.
+Another disadvantage of embedding the code within the sheet itself is that it prohibits versioning and sharing of the formulas separately from the data.
 
-However, spreadsheets have a number of downsides too, for example the lack of possibilities to abstraction.
-Instead of creating objects and methods on them, spreadsheet users mimic abstraction by copy-pasting formulas \todo{references naar LIVE paper}.
-Secondly, the editor that Excel and other spreadsheet systems provide is very limited, for example lacking syntax highlighting and not inviting proper structuring of formulas.
-
-\subsection{Code}
+%\subsection{Code}
 On the other hand, there is source code.
-Most existing programming languages do support abstraction, and modern editors help developers understand and structure their code.
-However, source code does not have liveness \todo{apart from a few recent initiative that have not been adopted mainstraim, Tijs, hedd gij dr paraat?}
+All existing programming languages support abstraction, and modern editors help developers understand and structure their code.
+However, mainstream integrated development environments (IDE) lack the live and interactive style of interaction so coveted by spreadsheet users.
+% \todo{apart from a few recent initiative that have not been adopted %mainstraim, Tijs, hedd gij dr paraat?}
 
@@
 
-\section{TrueGrid}
-In this paper, we will describe an approach bridging the gap between programming and spreadsheets and describe ts implications in both worlds.
-The basic concept of TrueGrid is a grid which can be programmed in a fully featured programming language, where the grid and the code are juxtapositioned ed, like in a spreadsheet, as demonstrated in Figure \ref{fig:TG}.
-Furthermore, like a spreadsheet, TrueGrid is live, i.e.
-on a change of data or code, the grid is updated.
+In this paper, we will describe \textit{TrueGrid}, a light-weight approach to bridging the gap between programming and spreadsheets and describe ts implications in both worlds.
+The basic concept of TrueGrid is to allow a spreadsheet like grid to be programmed using a full featured programming language, in a consistent user interface.
+Figure \ref{fig:TG} shows this idea as implemented in our prototype.
+The key charactertistic is that  developers see the code and data at once. Furthermore, like a spreadsheet, TrueGrid is live, i.e. on a change of data or code, the grid is updated.
+In our example, we use JavaScript as a language, however, the idea itself is not limited to a single programming language.
+% in fact, one can even image a grid in which different programming languages live.
 
@@
 
-In our example, we use JavaScript as a language, however, the idea itself is not limited to one programming language, in fact, one can even image a grid in which different programming languages live.
+In the next section we explore the implication of the TrueGrid user interface for spreadsheet users.
+Section~\ref{SECT:dev} explores TrueGrid from the perspective of programmers.
+In particular we discuss conventions for relating code to the grid view.
 
 \section{TrueGrid for Spreadsheet Users}
-As an implementation of TrueGrid for spredsheet users, we hypothesize that it will be embedded within an existing spreadsheet system, allowing users to use advanced features like PivotTables and charts on top of their TrueGrid.
+
+As an implementation of TrueGrid for spreadsheet users, we hypothesize that existing spreadsheet system is extended with a source code editor view, next to the grid-based view of the data.
+This allows users to  use advanced features like PivotTables and charts on top of their TrueGrid, but also use professional IDE services for expressing computations.
 For spreadsheet users, using TrueGrid over spreadsheets presents several benefits.
-For example, a programming language enables the use of modern editors.
-In the above example we have used Ace \todo{add link} providing syntax highlighting and syntax errors for free.
-Furthermore, the textual form of the code allows for easy diffing, merging enabling more mature version control on grids.
-While learning a new programming language can be challenging, there are spreadsheet developers working with VBA now, which is a fully featured language, and we envision TrueGrid having a lower threshold as the liveness and juxtaposition will ease understanding.
+For example, the use of a professional  editor supports the developer with editor services like syntax highlighting and error marking.
+Furthermore, the textual form of the code allows for easy diffing and merging, enabling more mature version control on spreadsheets.
+While learning a new programming language can be challenging, there are spreadsheet developers working with VBA now, which is a fully featured language.
+Unfortunately the integration between the code and a spreadsheet is low level and cumbersome.
+We envision TrueGrid having a lower lower learning curve because code and data are juxtaposed.
+
+\begin{figure}[t]
+\begin{minipage}{0.6\linewidth}
+\begin{lstlisting}
+function column_avg(row) {
+  return (row.lab + row.exam) / 2;
+}
+
+function cell_classAvg(col, row) {   
+  return sum(col) / col.length; 
+}
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.38\linewidth}
+\centering
+\sffamily
+\begin{tabular}{|c|c|c|c|}\hline
+student & lab & exam & avg\\\hline\hline
+Rich   & 7 & 8 & 7.5 \\\hline
+Jacome & 8 & 9 & 8.5\\\hline
+Birgit & 9 & 9 & 9 \\\hline
+ & & & \textit{classAvg:} 8.3 \\\hline
+\end{tabular}
+\end{minipage}
+\caption{Mockup of TrueGrid for spreadsheet use}
+\label{FIG:grades}
+\end{figure}
+
+As an example consider the simple grade book sheet shown in Fig.~\ref{FIG:grades}.
+It shows the actual data (both provided and computed) in the grid.
+The average and class average are expressed using ordinary Javascript functions.
+The TrueGrid environment links functions or methods to the cells in the grid using a naming convention.
+For instance, function starting with the \lstinline{column_} prefix computes a complete column, given a particular row (an ordinary Javascript object).
+This is used in the computation of the \lstinline{avg} column, where each cell contains the average of the lab and exam cells.
+The model also allows naming individual cells.
+This is illustrated in the function \lstinline{cell_classAvg}.
+The function receives the current column (\lstinline{col}, a Javascript array), and the current \lstinline{row}.
+Based on the elements in the column the class average is computed.
+
+One could imagine linking the code to the grid using row and column coordinates, just like ordinary spreadsheet formulas refer to (ranges of) rows and columns.
+However, this would lead to code that is not very intuitive if read separately.
+Furthermore,  insertion of rows and columns in the grid would require updating the source code itself, similar to how spreadsheet systems realign cell coordinates.
 
 \section{TrueGrid for Developers}
-Like spreadsheet users, developers often work with (example) data when programming.
-In modern IDEs, it is not possible to load data and then program alongside it.
-This makes it difficult for quickly prototype simple data analysis projects.
-A TrueGrid embedded in the IDE
+\label{SECT:dev}
 
-Using TrueGrid, a programmer you can inspect data and play with an initial implementation, before having to write interfaces or even unit tests.
-We envision this is a useful tool very early in the development process, when you have some data and only a vague idea of where the project is going.
+%In this section we switch perspective, and explore how TrueGrid can be beneficial for  professional programmers as well. 
+Like spreadsheet users, developers often work with (example) data when programming. For instance, developers use read-eval-print-loops (REPLs) to explore the behavior of a function. 
+Another use case is developer testing which requires setting up test fixtures and inspecting the results.
+Both these use cases, however, suffer from a lack of immediacy.
+After a change to the code, the developer needs to reenter expressions in the REPL to observe expected changes in behavior.
+Similar for testing: reexecuting the test is an explicit step after every change to the code.
+TrueGrid  eliminates these hickups and promises a more fluent, live experience.
 
-For example, consider the situation where a developer has a list of number that you want to manipulate \todo{We need a better example here} Before she even wants to write tests, the developer might want to explore the dataset a bit, for example, by sorting, calculating the mean or outliers.
-These type of activities currently have to be done outside of the IDE, for example in Excel or using a scripting language like R.
-Using TrueGrid this activity can be performed within the IDE.
+\begin{figure}[t]
+\begin{minipage}{0.6\linewidth}
+\begin{lstlisting}
+function leftpad(str, len, ch) {
+  str = String(str);
+  var i = -1;
+  if (!ch && ch !== 0) ch = ' ';
+  len = len - str.length;
+  while (++i < len) {
+    str = ch + str;
+  }
+  return str;
+}
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}{0.4\linewidth}
+\centering
+\sffamily
+\begin{tabular}{|c|c|m{0.5cm}|p{0.5cm}|c|}\hline
+function & 0 & 1 & 2 & Result \\\hline\hline
+leftpad & "foo" & 5 &  & "  foo" \\\hline
+leftpad & "foobar" & 6 & & "foobar" \\\hline
+leftpad & 1 & 2 & 3 & "01"\\\hline
+\end{tabular}
+\end{minipage}
+\caption{Exploring function behavior using TrueGrid}
+\label{FIG:leftpad}
+\end{figure}
 
-\section{Related Work}
-\todo{tsja}
+In particular, TrueGrid can be seen as a persistent REPL, where expressions or method invocations are continuously evaluated, after every change to the code or input data. 
+Consider the example shown in Fig.~\ref{FIG:leftpad}.
+On the left is a simple Javascript function for left-padding values with spaces or other padding characters. 
+Using TrueGrid, the programmer can provide example data in the grid and explore the implementation.
+This can be especially valuable early in the development process when  you have some data and only a vague idea of where the how certain functionality should be implemented.
+
+From the testing perspective, TrueGrid provides a kind of ``FIT testing on steroids''~\cite{mugridge2005fit}, where the grid functions as a live dashboard of test success and failure. 
+In this case, the grid shown in Fig.~\ref{FIG:leftpad} could have an additional column indicating the success or failure of the function execution, or use green/red coloring of rows to the same effect.
+Again, the success indicators would be automatically updated upon changing the source code on left.
+
+\section{Conclusion and Outlook}
+
+Spreadsheets are live programming environments.
+However, their lack of abstraction mechanisms and editor support are impediments to professional spreadsheet development.
+Conversely, traditional programming environments lack the continuos feedback that make spreadsheets so attractive.
+In this paper we have presented TrueGrid, a user interface design combining code editing and grid-based data view, with a live execution model.
+
+TrueGrid presents a promising bridge between the domain of spreadsheets and software development. It has the potential to improve end-user programming experience from two perspectives:
+\begin{itemize}
+\item For spreadsheet users: computations are expressed using program code, instead of  formulas in cells, so that end-users may enjoy both abstraction and liveness at the same time.
+\item For professional developers: spreadsheet-like grids support live exploration and testing of code, without explicitly invoking test scripts of entering expressions in a REPL.
+\end{itemize}
+We have implemented a prototype of TrueGrid which runs in the browser, using Javascript as the programming language.
+Further research is needed to empirically investigate the benefits of TrueGrid from both the programmer and spreadsheet user perspectives.
+We expect that TrueGrid provides a fruitful vehicle for exploring the middle ground between full end-user programming, and professional software development.
 

### 2016-04-19 true-grid/paper_SEMS16.tex
Fixed leftpad

@@
 leftpad & "foobar" & 6 & & "foobar" \\\hline
-leftpad & 1 & 2 & 3 & "01"\\\hline
+leftpad & 1 & 2 & 0 & "01"\\\hline
 \end{tabular}

### 2016-04-22 true-grid/paper_SEMS16.tex
updated screenshot
updated references and changed screenshot

@@
 %\subsection{Spreadsheets}
-Spreadsheets are very popular tools for end-user programming \todo{cite proof}.
+Spreadsheets are very popular tools for end-user programming.
 Their formula language is easy to learn and their grid interface is inviting.
@@
 When a users enters a formula and presses enter, they see the result, without long edit-compile-run cycles.
-This live characteristic of spreadsheets is often praised as their key success factor~\cite{proof}.
+This live characteristic of spreadsheets is often praised as their key success factor~\cite{HermansPhD}.
 \begin{figure}
@@
   \begin{center}
-  \includegraphics[width=9cm]{fig/TG.png}
-  \caption{A True Grid \todo{make ss with syntax highlighting, and add explanation}}
+  \includegraphics[width=12cm]{fig/TG.png}
+  \caption{A True Grid implementation, with the code editor on the left hand side and the grid on the righ, available via http://www.felienne.com/TrueGrid/}
   \label{fig:TG}

### 2016-06-16 true-grid/paper_SEMS16.tex
comments by reviewers addressed

@@
 In this paper we propose a method to bring both worlds closer together, by juxtaposing ordinary code and spreadsheet-like grids in the IDE, called TrueGrid.
-Using TrueGrid spreadsheet cells can be programmed using full featured programming language.
+Using TrueGrid spreadsheet cells can be programmed with a fully featured programming language.
 Spreadsheet users then may enjoy benefits of source code, including added abstractions, syntax highlighting, version control, etc.
 On the other hand, programmers may leverage the grid for interactive exploring and testing of code.
-We illustrate these benefits using a prototype implementation of TrueGrid that runs in the browser an uses Javascript as a programming language.
+We illustrate these benefits using a prototype implementation of TrueGrid that runs in the browser and uses Javascript as a programming language.
 \end{abstract}
@@
 This liveness is one of the core features of most spreadsheet systems.
-When a users enters a formula and presses enter, they see the result, without long edit-compile-run cycles.
+When a users enters a formula and presses enter, they see the result, without a lengthy edit-compile-run cycle.
 This live characteristic of spreadsheets is often praised as their key success factor~\cite{HermansPhD}.
@@
 Figure \ref{fig:TG} shows this idea as implemented in our prototype.
-The key charactertistic is that  developers see the code and data at once. Furthermore, like a spreadsheet, TrueGrid is live, i.e. on a change of data or code, the grid is updated.
+The key charactertistic is that  developers see the code and data at the same time. Furthermore, like a spreadsheet, TrueGrid is live, i.e. on a change of data or code, the grid is updated.
 In our example, we use JavaScript as a language, however, the idea itself is not limited to a single programming language.
@@
 
-As an implementation of TrueGrid for spreadsheet users, we hypothesize that existing spreadsheet system is extended with a source code editor view, next to the grid-based view of the data.
+As an implementation of TrueGrid for spreadsheet users, we hypothesize that existing spreadsheet system could be extended with a source code editor view, next to the grid-based view of the data.
 This allows users to  use advanced features like PivotTables and charts on top of their TrueGrid, but also use professional IDE services for expressing computations.
@@
 Unfortunately the integration between the code and a spreadsheet is low level and cumbersome.
-We envision TrueGrid having a lower lower learning curve because code and data are juxtaposed.
+We envision TrueGrid having a less steep learning curve, because code and data are juxtaposed.
 
@@
 The TrueGrid environment links functions or methods to the cells in the grid using a naming convention.
-For instance, function starting with the \lstinline{column_} prefix computes a complete column, given a particular row (an ordinary Javascript object).
+For instance, a function starting with the \lstinline{column_} prefix computes a complete column, given a particular row (an ordinary Javascript object).
 This is used in the computation of the \lstinline{avg} column, where each cell contains the average of the lab and exam cells.
@@
 Using TrueGrid, the programmer can provide example data in the grid and explore the implementation.
-This can be especially valuable early in the development process when  you have some data and only a vague idea of where the how certain functionality should be implemented.
+This can be especially valuable early in the development process when  you have some data and only a vague idea of how certain functionality should be implemented.
 
@@
 However, their lack of abstraction mechanisms and editor support are impediments to professional spreadsheet development.
-Conversely, traditional programming environments lack the continuos feedback that make spreadsheets so attractive.
+Conversely, traditional programming environments lack the continuous feedback that makes spreadsheets so attractive.
 In this paper we have presented TrueGrid, a user interface design combining code editing and grid-based data view, with a live execution model.
@@
 Further research is needed to empirically investigate the benefits of TrueGrid from both the programmer and spreadsheet user perspectives.
-We expect that TrueGrid provides a fruitful vehicle for exploring the middle ground between full end-user programming, and professional software development.
+We expect that TrueGrid provides a fruitful vehicle for exploring the middle ground between end-user programming on the one hand side, and professional software development on the other.
 

### 2016-06-17 true-grid/paper_SEMS16.tex
Related work

@@
 
+\section{Related Work}
+
+This work is positioned at the intersection of end-user programming and professional software development~\cite{EUSE}.
+The particular link between software engineering and spreadsheet has been explored before.
+For instance, Cunha et al.
+present model-based programming environments for spreadsheets~\cite{cunha2014model}.
+The focus of this work is to improve reliability of spreadsheet engineering by applying model-driven techniques. 
+In this work we primarily focus on improving programmer experience.
+
+Integrating spreadsheet-based user interfaces with code editors is a simple form of a heterogeneous programming environment (e.g.,~\cite{schrage2008beyond}), originally pioneered in the work on structure editors, and recently popularized by the Jetbrains Meta Programming System~\cite{MPS}.
+Although this kind of work aims for a much more invasive integration, it is in line with the goal of having best user interfaces for each aspect of the code.
+A similar strand of research is explored in the context of DSLs by Adam and Schultz~\cite{adam2015towards}.
+
 \section{Conclusion and Outlook}

### 2016-06-17 true-grid/paper_SEMS16.tex
Typos

--- 2016-06-16 true-grid/paper_SEMS16.tex
@@
 
+\section{Related Work}
+
+This work is positioned at the intersection of end-user programming and professional software development~\cite{EUSE}.
+The particular link between software engineering and spreadsheets has been explored before.
+For instance, Cunha et al.
+present model-based programming environments for spreadsheets~\cite{cunha2014model}.
+The focus of this work is to improve reliability of spreadsheet engineering by applying model-driven techniques. 
+In this work we primarily focus on improving programmer experience.
+
+Integrating spreadsheet-based user interfaces with code editors leads to a simple form of a heterogeneous programming environment (e.g.,~\cite{schrage2008beyond}), originally pioneered in the work on structure editors, and recently popularized by the Jetbrains Meta Programming System~\cite{MPS}.
+Although this kind of work aims for a much more invasive integration, it is in line with the goal of having the best user interface for each aspect of the programming experience.
+A similar strand of research is explored in the context of DSLs by Adam and Schultz~\cite{adam2015towards}.
+
 \section{Conclusion and Outlook}
